---
title: "dplyr_v2"
author: "Chris Kozak"
date: "December 1, 2017"
output: html_document
---

Dplyr is one of the most widely used packages in R. Developed by Hadley Wickham, it allows users to easily query data with a syntax that is similar to SQL. In this post, I will go through the basics of  dplyr queries using data from Major League Baseball and the NFL. First, we'll load the three packages we'll need for this exercise.

```{r message=FALSE}
library(nflscrapR)
library(dplyr)
library(Lahman)
```

##select and filter

Suppose we want to see the home run totals for the 2016 New York Yankees. We can easily do this using the **select** and **filter** commands in dplyr.

```{r}
Batting%>%
  select(playerID,yearID,teamID,HR)%>%
  filter(teamID=="NYA" & yearID==2016)
```

Note that we're using two equals signs instead of just one. In dplyr -- and most query languages -- a single equals designates a new variable, while double equals indicates equality. For the next exercise, let's say we want to select players who had more than 40 home runs, but less than 60 strikeouts. This would indicate a great batter who rarely strikes out -- making him a real asset to the team. For this task, we'll use the greater than (>) and less than (<) signs.

```{r}
Batting%>%
  select(playerID,yearID,HR,teamID,SO)%>%
  filter(HR > 40 & SO < 60)
```

For the next exercise, let's say we want to find all the Phillies players who hit over 30 homeruns during the 1970s. For this, we'll use a combination of specific operators:

```{r}
Batting%>%
  select(playerID,yearID,HR,teamID)%>%
  filter(teamID == "PHI" & yearID <=1979 & yearID>=1970 & HR>30)
```

Notice that in this case, dplyr gives us that data we wanted, but let's say we want it in a specific order -- with the highest number of home runs at the top. For that, we would use the arrange command.


##arrange

Let's run a similar command next. We'll select players from *all* teams this time, but filter out the ones who hit less than 51 home runs, and we'll arrange the data with the most home runs at the top. For good measure, we'll also throw in strikeouts, yearID, and at-bats to get a more holistic view of each player.

```{r}
Batting%>%
  select(playerID,yearID,teamID,HR,SO,AB)%>%
  filter(HR>50)%>%
  arrange(desc(HR))
```

Notice that we used the desc() argument with homeruns to ensure the highest number of home runs is at the top of the table. Next, we'll figure out which players have a ton of at-bats and almost no strikeouts.

```{r}
Batting%>%
  select(playerID,yearID,HR,teamID,SO,AB)%>%
  filter(AB>=400 & SO<10)%>%
  arrange(SO)
```

In this case, we left out the desc() argument, since we want the best players -- those with the least strikeouts -- at the top.


##group_by and summarize

These two commands allow us to easily take all of the records for one observation and summarize them. For example, we want to see how many touchdowns Aaron Rodgers made in 2016 using the nflscrapR package.

```{r}
playerstats16%>%
  select(name,date,pass.tds,rushtds)%>%
  filter(name=="T.Brady")
```

We get 12 records, representing one record for each game that Tom Brady played in during the 2016 regular season. (He missed the first four due to the Deflategate suspension.) Now, we'll use the group_by() and summarize() commands to put all of his records together for the season:

```{r}
playerstats16%>%
  group_by(name) %>% 
  select(name,pass.tds,rushtds) %>% 
  summarize(passing_tds=sum(pass.tds),
            rush_tds=sum(rushtds)) %>% 
  filter(name=="T.Brady")
```
Now, we can see that Tom Brady had 28 touchdowns for the season, and all by passing. Next, let's take these same principles to see who had the most reception yards on the Steelers for the 2016 season.

```{r}
playerstats16%>%
  group_by(name) %>% 
  select(name,Team,recyds)%>%
  filter(Team=="PIT" & recyds > 0) %>% 
  summarize(season_recyds=sum(recyds)) %>% 
  arrange(desc(season_recyds))
```

This shows, to little surprise, that Antonio Brown had the most receptions yards by far, followed by Le'Veon Bell and Eli Rodgers.


##mutate

Finally, we'll review the mutate() command, which can compute and append a new column. With base R, we would first create a new column of data, then attach it to our data set. Let's say we want to create a new column featuring each QB's total TD/INT ratio from 2016. The playerstats16 data set from nflscrapR does not have a column specifically for this, but we can easily add it with the mutate command.

```{r}
playerstats16%>%
  group_by(name) %>% 
  select(name,pass.tds,rushtds,pass.ints)%>%
  summarize(season_tds=sum(pass.tds+rushtds),
            season_ints=sum(pass.ints)) %>%
  mutate(td.int=season_tds/season_ints) %>% 
  filter(td.int > 0 & td.int != "Inf") %>% 
  arrange(desc(season_tds))
```

Here, we can see that Aaron Rodgers threw (or ran) about 6 touchdowns for every one interception he threw. Phillip Rivers, of the then-San Diego Chargers, only made 1.5 touchdowns per each interception. Tom Brady had one of the strongest ratios, making 14 touchdowns per every one interception.
